---
title: 'R tutorial Peru: trait data exploration'
author: "Julia Chacón"
date: "22/3/2020"
output: html_document
---

<style>
p.caption {
  font-size: 0.8em;
}
</style>

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r load-packages, include=FALSE}
library("tidyverse")
library("dataDownloader")
library("vegan")
library("ggvegan")
library("patchwork")
library("glue")
#library("ggfortify")
```

```{r read trait data, include=FALSE}
traits <- read_csv(file = "traits/data/PFTC3.7_Traits_2018_Peru_cleaned.csv")
```

# PFTC3 Wayqecha (PERÚ)

## Exploring how the data are organized

### Trait data

We suggest starting exploring the data sets of traits, asking for the names of the columns, the levels of each factor, and dimensions.

```{r Trait data, eval=FALSE}
colnames(traits)
```

Explore a bit the levels of each factor, and the variable names. 

```{r Trait data2, eval=TRUE}
unique(traits$Elevation)
```

```{r Trait data3, eval=TRUE}
unique(traits$Treatment)
unique(traits$Site)## just change here the name of the column you want to explore
```

## Multivariate exploratory
### PCA

This document explains how to perform a PCA using `ggfortify` package and `ggplot::autoplot`

First, subset the big "traits" tibble into a smaller one. Log-transform some of the traits, select a few traits that are quite complete and the columns Taxon and Site.

```{r transform-data, eval=TRUE}
traits.sel <- traits %>% # trait selection
    mutate(Leaf_Thickness_Ave_mm.log = log(Leaf_Thickness_Ave_mm), 
           Leaf_Area_cm2.log = log(Leaf_Area_cm2)) %>%
    select(Taxon, Site, Plant_Height_cm,
           Leaf_Thickness_Ave_mm.log, Leaf_Area_cm2.log)
```

Select only those data from the site Wayquecha. For doing the PCA we only need the traits values. Eliminate the NAs too.

```{r filter-data, eval=TRUE}
traits.WAY <- traits.sel %>%
  filter(Site == "WAY") %>%
  drop_na()
```

Now, store the result of the PCA into an object pca.WAY. Eliminate the columns Site and Taxon, we don't need them for the PCA. 
Then, do the PCA using `vegan::rda()`. 
We use the argument `scale = TRUE` so that each variable is transformed to have a mean of zero and unit variance. 
This is necessary because the variables have different scales.

```{r PCA, eval=TRUE}
pca.WAY <- traits.WAY %>% # store result as "pca.WAY"
  select(-Site, -Taxon) %>%   # remove Site & Taxon columns
  rda(scale = TRUE) #do the PCA
```

And plot it using the `autoplot` function (this is coming from the `ggvegan` package, and is using `autoplot.rda`). 

```{r plot-PCA, eval=TRUE}
autoplot(pca.WAY) +
  theme(legend.position = "none")
```

Every leaf appears as a red dot. 
The traits are shown by arrows. Unfortunately the defaults have made the arrows too long, but we can control that with the `const` argument to `autoplot.rda` (You will need the latest version of `ggvegan` for this to work - install with `remotes::install_github("richardjtelford/ggvegan")`).

```{r better-plot-PCA, eval=TRUE}
autoplot(pca.WAY, const = c(0.2, 1)) + 
  xlim(-.19, NA) + # make more space for the labels
  theme(legend.position = "none")
```



It would be nice to show PCA with different colours for the different species.
We cannot do that directly with `autoplot.rda()`, but we can do it manually, using `fortify()` to extract the scores from the PCA.
This gives us lots of control over the plot.

```{r better-plot-PCA2, eval=TRUE}
#extract scores from ordination
fpca.WAY <- fortify(pca.WAY, const = c(0.2, 1), axes = 1:2)

#get the site scores (ie each leaf)
fpca.WAY_leaves <- fpca.WAY %>% 
  filter(Score == "sites") %>% 
  bind_cols(traits.WAY)

# get the trait/variable scores (species in vegan terminology)
fpca.WAY_traits <- fpca.WAY %>% 
  filter(Score == "species") 


#plot the leaf scores         
g <- ggplot(fpca.WAY_leaves, aes(x = PC1, y = PC2)) +
  geom_point(aes(colour = Taxon), show.legend = FALSE) +
  coord_equal() # force scale to be correct

#add trait arrows to plot
g <- g +
  geom_segment(data = fpca.WAY_traits,
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.2, "cm")),
               colour = "navy") +
  geom_text(data = fpca.WAY_traits, # add labels
            aes(x = PC1, y = PC2, label = Label), nudge_x = -0.005, size = 3, hjust = 1)

g <- g + 
  xlim(-.20, NA) + #space for labels
  theme(legend.position = "none")# no legend as too large
g
```

We can finish off the plot by adding the proportion of variance explained by each axis to the plot. 
We can calculate this with the function `eigenvals`. 
We can make the labels with `glue::glue()` or with `paste()`.  

```{r better-plot-PCA3}
explained <- eigenvals(pca.WAY)[1:2] / sum(eigenvals(pca.WAY)) * 100
explained <- round(explained, 1)

g + 
  labs(x = glue("PC1 [{explained[1]}%]"), y = glue("PC1 [{explained[2]}%]")) 
```

