---
title: 'R tutorial Peru: trait data exploration'
author: "Julia Chacón"
date: "22/3/2020"
output: html_document
---

<style>
p.caption {
  font-size: 0.8em;
}
</style>

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r Call the libraries, include=FALSE}
library("tidyverse")
library("dataDownloader")
library("DBI")
library("vegan")
library("ggvegan")
library("patchwork")
library("ggfortify")
```

```{r read trait data, include=FALSE}
traits <- read_csv(file = "traits/data/PFTC3.7_Traits_2018_Peru_cleaned.csv")
```

# PFTC3 Wayqecha (PERÚ)

## Exploring how the data are organized

### Trait data

We suggest starting exploring the data sets of traits, asking for the names of the columns, the levels of each factor, and dimensions.

```{r Trait data, eval=FALSE}
colnames(traits)
```

Explore a bit the levels of each factor, and the variable names. 

```{r Trait data2, eval=TRUE}
unique(traits$Elevation)
```

```{r Trait data3, eval=TRUE}
unique(traits$Treatment)
unique(traits$Site)## just change here the name of the column you want to explore
```

## Multivariate exploratory
### PCA

This document explains how to perform a PCA using "ggfortify" package and "ggplot: autoplot"

First, subset the big "traits" tibble into a smaller one. Log-transform some of the traits, select a few traits that are quite complete and the columns Taxon and Site.

```{r PCA, eval=TRUE}
traits.sel <- traits %>% # trait selection
    mutate(Leaf_Thickness_Ave_mm.log = log(Leaf_Thickness_Ave_mm), 
             Leaf_Area_cm2.log = log(Leaf_Area_cm2)) %>%
                 select(Taxon, Site, Plant_Height_cm,
                     Leaf_Thickness_Ave_mm.log, Leaf_Area_cm2.log)
```

Select only those data from the site Wayquecha. For doing the PCA we only need the traits values. Eliminate the NAs too.

```{r PCA2, eval=TRUE}
traits.WAY <- traits.sel %>%
  filter(Site=="WAY") %>%
    drop_na()
```

Now, store the result of the "pca" into an object pca.WAY. Eliminate the columns SIte and Taxon, we don´t need them for the pca. For that we first scale to mean 0 and unit variance. Then, do the pca using prcomp()

```{r PCA3, eval=TRUE}
pca.WAY <- traits.WAY %>% # store result as "pca"
  select(-Taxon) %>%   # remove Taxon column
  select(-Site) %>% # remove Site column
  scale() %>%            # scale to 0 mean and unit variance
  prcomp() #do the pca
```

And plot it using "autoplot" function. 
```{r PCA4, eval=TRUE}
autoplot(pca.WAY)
```

Every species appeared as a black dot. We can colour that based on the Taxon name (species name). So, every taxon is going to be plotted in a different colour.

Better plot it without the legend, because there are too many species and the legend is huge. But you are free of plotting the legend if you want. 

```{r PCA5, eval=TRUE}
autoplot(pca.WAY, data = traits.WAY, colour = 'Taxon') +
  theme(legend.position = "none") # without a legend
```

You can plot the loadings on top:
```{r PCA6, eval=TRUE}
autoplot(pca.WAY, data = traits.WAY, colour = 'Taxon', loadings = TRUE) +
  theme(legend.position = "none")
```

Even play with the loadings colour, labels and size. Again, witout the legend

```{r PCA7, eval=TRUE}
autoplot(pca.WAY, data = traits.WAY, colour = 'Taxon',
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3)+
  theme(legend.position = "none")
```
